---
title: "QueryingGO"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GO terms to gene list
Sometimes you want to get a list of all the genes that are associated with a specific GO term (or multiple GO terms).
This is super easy. Just copy and paste the code snippits into RStudio

The first few lines just make sure you have the proper software installed. (You can comment out )

```{r biomart, eval=F}
source("http://bioconductor.org/biocLite.R")
#install package if not already installed
biocLite("biomaRt")
```
Hopefully everything is k after that.

Now a good way to find GO terms of interest is to mess around on http://www.ebi.ac.uk/QuickGO
Let's say we have a dope term GO:0030182, Neuron Differentiation. We can make a list of all the genes that are associated with this GO term
```{r one}
#load the package
library(biomaRt)
#This is your list of terms
termList <- c("GO:0030182")
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host="www.ensembl.org")

GeneData <- getBM(attributes=c('hgnc_symbol',"name_1006"),
                   filters = c('go'), values = termList, mart = mart)

print(head(GeneData))

``` 


We could also do this with a list of terms:
```{r many}
termList <- c("GO:0007399","GO:0097458","GO:0045202")
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host="www.ensembl.org")

GeneData <- getBM(attributes=c('hgnc_symbol'),
                   filters = c('go'), values = termList, mart = mart)

print(head(GeneData))

``` 
attributes is the list of things you want back from the database, in this case the gene symbol 'hgnc_symbol' . You can string multiple together by separating them with commas (You might want to include 'name_1006' for the names of all the terms associated with the genes as well)
filters is the data type that you're sending to the database, in this case the GO ID, 'go'.
values is the list of items you're querying

###Now save the list to a file on your desktop
This will write the data stored as "GeneData" to a file on your desktop
```{r saves}
write.table(GeneData,"~/Desktop/GO_Output.txt",sep = "\t",row.names = F,col.names = F)
```

That's all you have to do to get a list of genes from a list of GO terms!